\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-05-30 eg6-letters/NIM/EG6_RTPC_NIM.tex
some cleaning
Few minor edits

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at Jefferson 
Laboratory to track low-energy nuclear recoils for the purpose of measuring
exclusive nuclear channels, such as coherent Deeply Virtual Compton Scattering
and coherent meson production on $^4$He. In such processes, the $^4$He nucleus
remains intact and recoils as a whole in the final state. The CEBAF Large
Acceptance Spectrometer (CLAS) has been used for similar studies on proton
targets, however it cannot track low energy $\alpha$ particles. In
2009, we carried out measurements using the CLAS spectrometer supplemented by
the RTPC positioned directly around a gaseous $^4$He target, allowing a detection
threshold as low as $~$150  MeV/c for $^4$He. This article discuses the design,
work principle, calibration methods and performances of the RTPC.
\end{abstract}

@@

Thomas Jefferson National Accelerator Facility (TJNAF), in Newport News, Virginia,
USA, provides high power electron beams of energy up to 6 GeV and 100$\%$ duty cycle
to three experimental Halls (A, B, C) simultaneously. The CLAS spectrometer, located
in Hall-B, is composed of several sub-detectors and two magnets. The spectrometer
was designed to track charged particles of momentum greater than 250 MeV/c as well 
as detecting real photons. Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) for the tracking of charged 
    particles~\cite{DCref} (in blue on the figure).
 \item Superconducting toroidal magnet (in yellow) to bend the trajectories 
    of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Cherenvkov Counters (CC) to distinguish electrons from other negative 
    particles, such as pions, at lower energies~\cite{CCref} (in pink).
 \item Scintillation Counters (SC) to identify hadrons by measuring their 
    time of flight (TOF)~\cite{TOFref} (in red).
 \item Electromagnetic Calorimeters (EC) to distinguish electrons from other negative 
    particles at higher energies and to detect neutral particles (neutrons and 
    photons in particular)~\cite{ECref} (in green).
\end{itemize}

@@

Measurement of the Deeply Virtual Compton Scattering (DVCS) process
($eH \rightarrow e' H' \gamma$, where $H$ is a hadron) necessitates an upgrade
of this setup.  With a 6 GeV electron beam, the majority of DVCS photons are produced
at very forward angles, where the acceptance of the EC is poor. To extend the detection range,
an inner calorimeter (IC) was built for the E01-113 experiment in 2005.
The IC is constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$. Each crystal is 16 cm long, corresponding
to 17 radiation lengths. The achieved energy resolution is around 3 to 4$\%$ for photon
energies between 2 GeV and 5 GeV, with an angular resolution between 3 and 5 mrad 
\cite{Hyon-suk}. To protect this detector from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the IC. 
In addition, the recoil nuclei of the coherent DVCS on helium was detected 
using a RTPC developed to track low energy nuclear fragments. The solenoid field
serving as the analyzing magnet for the particle tracking in the RTPC. This 
setup was used during a three months experimental run
in 2009 with a longitudinally polarized electron beam of 130 nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

In this paper we report the design, calibration, and performance of the radial 
TPC, organized as follows. In section \ref{sec_design}, we detail the design 
and internal structure of the detector. Section \ref{sec_readout} describes 
the properties of the read-out system. Calibration 
strategies are discussed in section \ref{sec_calib}. Finally, the performances
of the RTPC are described in section \ref{sec_perf}.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from coherent 
DVCS have an average momentum around 250 MeV/c. Such low energy $\alpha$ 
particles are stopped very rapidly, so the RTPC was designed to be as close 
as possible to the target and fitting within the 230~mm diameter 
of the 5 Tesla solenoid magnet surrounding the target \cite{Hyon-suk}.

@@

We detail here the different regions shown in figure~\ref{fig:RTPC_1_4} starting 
from the beam line and towards larger radius:\\
\begin{itemize}
   \item First, the 6 atm helium-4 target extends along the beam line axis, it 
      is 284~mm long with a 3~mm radius wall made of a 27-$\mu$m-thick Kapton.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o ller electrons scattered by the beam. This 
      region is surrounded by a 4 $\mu$m thick window made of aluminized Mylar 
      and electronically grounded.
   \item The second gap region extends between 20~mm and 30~mm and is filled with a 
      gas mixture of 80$\%$ neon and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which 
      serves as the cathode and is set at 4260~V.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first Gas Electron Multiplier(GEM), 60 mm away 
      from the beam axis. The electric field in this region is about 550~V/cm 
      and perpendicular to the beam.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied to each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Preamplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

\begin{figure}[tbp]
\centering
\includegraphics[scale=0.70]{fig/GEM_photo.jpg}
\caption{Picture of a GEM foil similar to the one used for our RTPC.} 
\label{fig:GEMs}
\end{figure}

The GEM technology has been chosen for this RTPC for its flexibility, 
allowing a curved amplification surface. Also, GEMs are known to rarely create 
sparks~\cite{GEM_ref}, which is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 50~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in figure~\ref{fig:GEMs}. A potential difference of 275 V 
is applied between the two 
copper layers of the GEM, creating a very strong electric field in each hole 
leading to high ionization and therefore amplification of the incoming 
electrons. A high potential of about 1600 V/cm is set between 
the GEMs to push the electrons toward the next GEM and eventually the readout 
pads. \\

The drift gas used in the experiment is a 80-20\% neon-DME mixture, this choice 
has been made in order to balance the energy deposit with a reasonnable
Lorentz angle. Calculations using the MAGBOLTZ program~\cite{MAGBOLTZ} 
showed that with the perpendicular 5~T field, we would have a Lorentz angle of 
about 23$^\circ$ with this mixture. This is already significant and a heavier
gas would have larger angle increasing significantly
the length of the drift paths as the electron would 
make a long spiral around the beam line before to reach the GEMs.

One of the important problematic to develop the radial TPC was to obtain a 
good support structure of the GEM foils, allowing an easy installation of the 
GEM in good conditions and with high precision. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfuly realized these prerequisites 
by using fiber glass rings glued to each ends of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
all gluing and soldering operations.  Because the rigidity of the high 
rigidity of the GEM foil, the structure was selfsupporting and only the 
backward part of the cylinder was fixed to the main mechanical structure. This 
design only left a light fiberglass ring in the forward region, reducing to a 
minimum secondary interactions.

@@

{\bf \color{red} The use of the term TDC is improper and should be removed
everywhere. Replace by T.}

@@

\bibitem{proposal}
   K.Hafidi et al., Deeply virtual Comton scattering off $^{4}$He, Jlab 
   proposal to PAC33 (2007).

@@

>>>>

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at Jefferson 
Laboratory to track low-energy nuclear recoils for the purpose of measuring
exclusive nuclear channels, such as coherent Deeply Virtual Compton Scattering
and coherent meson production on $^4$He. In such processes, the $^4$He nucleus
remains intact in the final state, however the CEBAF Large
Acceptance Spectrometer (CLAS) cannot track the low energy $\alpha$ particles. In
2009, we carried out measurements using the CLAS spectrometer supplemented by
the RTPC positioned directly around a gaseous $^4$He target, allowing a detection
threshold as low as 300~MeV/c for $^4$He. This article discuses the design,
work principle, calibration methods and performances of this RTPC.
\end{abstract}

@@

Thomas Jefferson National Accelerator Facility (TJNAF), in Newport News, Virginia,
USA, provides high power electron beams of energy up to 6 GeV and 100$\%$ duty cycle
to three experimental Halls (A, B, C) simultaneously. The CLAS 
spectrometer~\cite{CLASref}, located
in Hall-B, is composed of several sub-detectors and two magnets. 
Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) for the tracking of charged 
    particles~\cite{DCref} (in blue on the figure).
 \item Superconducting toroidal magnet (in yellow) to bend the trajectories 
    of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Cherenvkov Counters (CC) to distinguish electrons from other negative 
    particles, such as pions, at lower energies~\cite{CCref} (in pink).
 \item Scintillation Counters (SC) to identify hadrons by measuring their 
    time of flight (TOF)~\cite{TOFref} (in red).
 \item Electromagnetic Calorimeters (EC) to distinguish electrons from other negative 
    particles at higher energies and to detect neutral particles (neutrons and 
    photons in particular)~\cite{ECref} (in green).
\end{itemize}

@@

Measurement of the Deeply Virtual Compton Scattering (DVCS) process
($eH \rightarrow e' H' \gamma$, where $H$ is a hadron) necessitates an upgrade
of this setup.  With a 6 GeV electron beam, the majority of DVCS photons are produced
at very forward angles, where the acceptance of the EC is poor. To extend the detection range,
an inner calorimeter (IC) was built for the E01-113 experiment in 2005.
The IC is constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$. Each crystal is 16 cm long, corresponding
to 17 radiation lengths. The achieved energy resolution is around 3 to 4$\%$ for photon
energies between 2 GeV and 5 GeV, with an angular resolution between 3 and 5 mrad 
\cite{Hyon-suk}. To protect this detector from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the IC. 
In addition, the recoil nuclei of the coherent DVCS on helium was detected 
using a RTPC developed to track low energy nuclear fragments. The solenoid field
serving as the analyzing magnet for the particle tracking in the RTPC. This 
setup was used during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

In this paper we report the design, calibration, and performance of the radial 
TPC, organized as follows. In section \ref{sec_design}, we detail the design 
and internal structure of the detector. Section \ref{sec_readout} describes 
the properties of the read-out system. Calibration 
strategies are discussed in section \ref{sec_calib}. Finally, the performances
of the RTPC are described in section \ref{sec_perfor}.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from coherent 
DVCS have an average momentum around 250 MeV/c. Such low energy $\alpha$ 
particles are stopped very rapidly, so the RTPC was designed to be as close 
as possible to the target and fitting within the 230~mm diameter 
of the 5 Tesla solenoid magnet surrounding the target.

@@

We detail here the different regions shown in figure~\ref{fig:RTPC_1_4} starting 
from the beam line and towards larger radius:\\
\begin{itemize}
   \item First, the 6 atm helium-4 target extends along the beam line axis, it 
      is 284~mm long with a 3~mm radius wall made of a 27-$\mu$m-thick Kapton.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o ller electrons scattered by the beam. This 
      region is surrounded by a 4 $\mu$m thick window made of aluminized Mylar 
      is electronically grounded.
   \item The second gap region extends between 20~mm and 30~mm and is filled with a 
      gas mixture of 80$\%$ neon and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which 
      serves as the cathode and is set at 4260~V.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first Gas Electron Multiplier(GEM), 60 mm away 
      from the beam axis. The average electric field in this region is about 550~V/cm 
      and perpendicular to the beam.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied to each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Preamplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

\begin{figure}[tbp]
\centering
\includegraphics[scale=0.70]{fig/GEM_photo.jpg}
\caption{Picture of a GEM foil similar to the one used for our RTPC. {\color{red} Source?}} 
\label{fig:GEMs}
\end{figure}

The GEM technology has been chosenfor its flexibility as it
allows for a curved amplification surface. Also, GEMs are known to rarely create 
sparks~\cite{GEM_ref}, which is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 50~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in figure~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

The drift gas used in the experiment is a 80-20\% neon-DME mixture, this choice 
has been made in order to balance the energy deposit with a reasonnable
Lorentz angle. Calculations using the MAGBOLTZ program~\cite{MAGBOLTZ} 
showed that with the perpendicular 5~T field, we would have a Lorentz angle of 
about 23$^\circ$ with this gas mixture. This is already a significant angle,
a heavier gas would have larger angle increasing significantly
the length of the drift paths as the electron would 
make a long spiral around the beam line before to reach the GEMs.

One of the important problematic to develop the radial TPC was to obtain a 
good support structure of the GEM foils, allowing an easy installation of the 
GEM in good conditions and with high precision. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfuly realized these prerequisites 
by using fiber glass rings glued to each ends of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
all gluing and soldering operations. The rigidity of the GEM foil is enough
for the structure to be selfsupporting and only the 
backward part of the cylinder was fixed to the main mechanical structure. This 
design only left a light fiberglass ring in the forward region, reducing to a 
minimum secondary interactions.

@@

@@

@@

\bibitem{proposal1}
   G Asryan et al., Meson spectroscopy in the Coherent Production on $^{4}$He with CLAS, Jlab 
   proposal to PAC31 (2007).

\bibitem{proposal2}
   K. Hafidi et al., Deeply virtual Comton scattering off $^{4}$He, Jlab 
   proposal to PAC33 (2008).
